---
title: "Python to R practice"
output: pdf_document
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "build") })
---

```{r setup, echo=FALSE, include=FALSE}
# Setup
knitr::opts_chunk$set(results = 'hide', echo = FALSE, fig.keep = 'none', 
                      warning = FALSE, message = FALSE)

library(pacman)
pacman::p_load(tidyverse, reticulate, knitr, kableExtra)
use_condaenv("practice_env")
```

```{r generate-data, echo=FALSE, eval=TRUE}
set.seed(123)
n <- 1000

ids <- 1:n
zip_codes <- c(
  "98101", "98102", "98103", "98104", "98105", "98106", "98107", "98108", "98109", "98112",
  "98115", "98116", "98117", "98118", "98119", "98121", "98122", "98125", "98126", "98133",
  "98134", "98136", "98144", "98146", "98154", "98164", "98174", "98177", "98178", "98195",
  "98199"
)
ages <- sample(25:100, n, replace = TRUE)
smoking <- sample(0:1, n, replace = TRUE)
years_smoked <- sample(0:10, n, replace = TRUE)

data <- data.frame(
  id = ids,
  zip_code = sample(zip_codes, n, replace = TRUE),
  age = ages,
  smoking = smoking,
  years_smoked = ifelse(smoking == 1, sample(0:10, n, replace = TRUE), NA)
)

write.csv(data, "data.csv", row.names = FALSE)
```


## about the data

the data we generated is called `data.csv`. it has the following columns:

- `id`: unique identifier
- `zip_code`: seattle area ZIP code
- `age`: age in years
- `smoking`: binary variable (0 = does not smoke, 1 = smokes)
- `years_smoked`: number of years smoked (NA if non-smoker)

## python code to translate

```{python python-code, echo=TRUE, results='asis', eval=TRUE}
import pandas as pd
import numpy as np
import scipy.stats as stats

df = pd.read_csv("data.csv")

# loop over columns and print out means for age and smoking
for column in list(df.columns.values):
    if column in ['age', 'smoking']:
        colmean = np.mean(df[column])
        print("{col} mean is {mean}\n".format(col=column, mean=colmean))

# rename a variable
df.rename(columns={'age': 'age_in_years'}, inplace=True)

# subset based on a value
df = df[df['age_in_years'] > 50]

# create a new variable by adding or multiplying two variables
df['age_first_smoked'] = df['age_in_years'] - df['years_smoked']

# drop rows with missing values for the statistical test
df_test = df.dropna(subset=['age_first_smoked'])

# create categorical variables for chi-square test (chi-square requires categorical data)
# use quantiles to ensure balanced groups
df_test['age_group'] = pd.qcut(df_test['age_in_years'], q=3, labels=['50-62', '62-72', '72+'])
df_test['age_first_smoked_group'] = pd.qcut(df_test['age_first_smoked'], q=3, labels=['Low', 'Medium', 'High'])

# compute a summary statistic (chi-square test in this example)
crosstab = pd.crosstab(df_test['age_group'], df_test['age_first_smoked_group'])
chi2_stat, p_val, dof, expected = stats.chi2_contingency(crosstab)

print(f"chi-square test statistic: {chi2_stat}")
print(f"P-value: {p_val}")
print(f"degrees of freedom: {dof}")
```

## python output preview

the first 5 rows of the output from the python code:

```{python python-head, echo=TRUE, results='asis', eval=TRUE}
head = df.head()
print(head.to_markdown())
```

## your turn: write the r translation

write a code chunk below that does the same thing as the python code above. your code should:

1. read in the data
2. loop over columns and print means for `age` and `smoking`
3. rename `age` to `age_in_years`
4. subset to only include rows where `age_in_years > 50`
5. create a new variable `age_first_smoked` = `age_in_years` - `years_smoked`
6. drop rows with missing values in `age_first_smoked`
7. create categorical variables for age groups and age_first_smoked groups using quantiles (pd.qcut with q=3)
8. run a chi-square test and print the results

```{r your-answer}
# YOUR CODE HERE

```

\newpage

# appendix: all code

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```
